---
title: "STAT40830 - Homework 1"
author: "Akash Ajaykumar Nair - 24217873"
format: 
  pdf:
    echo: false
    geometry:                  
      - top = 0.5in
      - bottom = 0.5in
      - left = 0.5in
      - right = 0.5in
      - a4paper
    include-in-header: 
      text: |
        \usepackage{xcolor}
editor: visual
execute: 
  error: true                        
  warning: false
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
# Load required libraries
library(palmerpenguins)
library(ggplot2)
```

# Introduction: Penguin Species and Their Physical Traits

The **Palmer Penguins** dataset provides size measurements for three penguin species: \textcolor{red}{Adelie, Chinstrap and Gentoo} living in the Palmer Archipelago, Antarctica. It includes data on features like bill dimensions, flipper length and body mass, along with the species and island of each observation.
This report uses a scatter plot to examine how bill length and bill depth vary by species. These features are often used to distinguish between penguin types.

# Dataset Summary

Variables:

- `bill_length_mm`: Bill length in millimeters
- `bill_depth_mm`: Bill depth in millimeters
- `species`: Penguin species (*Adelie, Chinstrap, Gentoo*)

```{r}
# Summary of dataset
summary(penguins[, c("species", "bill_length_mm", "bill_depth_mm")])
```

# Scatter Plot: Bill Length vs Bill Depth

```{r, fig.align='center'}
# Create scatter plot
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 2) +
  labs(
    title = "Penguin Bill Dimensions by Species",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    color = "Species"
  ) +
  theme_minimal()
```

# Description of the plot 

This scatter plot shows clear differences in bill dimensions among species:

- Adelie penguins have shorter, deeper bills.

- Gentoo penguins have longer, shallower bills.

- Chinstrap penguins fall in between.

This separation makes bill measurements a useful tool for distinguishing species.

# Conclusion
The plot clearly demonstrates how bill dimensions can effectively distinguish penguin species, making these features useful in classification models or ecological field studies.